---
title: "Placeholder for article 3"
author:
  - Jane Doe:
      correspondence: "yes"
      email: janedoe@fosg.org
      orcid: 0000-0003-1689-0557
      institute:
        - fosg
        - fop
  - John Q. Doe:
      institute: fosg
      orcid: 0000-0003-1689-0558
  - Peder Ås:
      institute: fosg
      orcid: 0000-0003-1689-0559
  - Juan Pérez:
      orcid: 0000-0003-1689-0551
      institute:
        - name: Acme Corporation
  - Max Mustermann:
      orcid: 0000-0003-1689-0552
institute:
  - fosg:
      name: Formatting Open Science Group
      address: 23 Science Street, Eureka, Mississippi, USA
  - fop: Federation of Planets
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The actual document text starts here: -->

# Introduction

Here is a citation [@Marwick2017]

# Background

# Methods

# Results

<!-- Here's some example analysis code: -->

```{r}
#| label: library

library(readr)
library(dplyr)
library(tidyr)
library(stringr)

```


```{r}
#| label: get-data
#| eval: false

# Technological data
gravettian_tech <- read.csv(here::here('analysis/data/raw_data/VBT_6_7_JB.csv'))
proto_tech <- read.csv(here::here('analysis/data/raw_data/VBT_4E_5_JB.csv'))
solutrean_JC <- read.csv(here::here('analysis/data/raw_data/VBR_ABC_JC.csv'))
solutrean_points <- read.csv(here::here('analysis/data/raw_data/shelter_points_JC.csv'))

# RM data
shelter_RM <- read_csv(here::here('analysis/data/raw_data/MACRO_IND_SHELTER.csv')) # Shelter RM analysis
terrace_RM <- read_csv(here::here('analysis/data/raw_data/MACRO_IND_TERRACE.csv')) # Terrace RM analysis

# Context data
context <- read_csv(here::here('analysis/data/raw_data/VB_Terrace_context.csv')) # Terrace Field DB - lithic artefact info
xyz <- read_csv(here::here('analysis/data/raw_data/VB_Terrace_xyz.csv')) # Terrace Field DB - just coordinates

```

```{r}
#| label: terrace-databases-setup

cols_to_concat_terrace <- c("UNIT", "ID")
context$ID <- str_squish(context$ID) # Remove empty spaces from ID field
xyz$ID <- str_squish(xyz$ID) # Remove empty spaces from UNIT field

proto_tech_WIP <- proto_tech %>%
  tidyr::unite(col='ID', all_of(cols_to_concat_terrace), sep="-", remove=TRUE) %>% 
  rename(CLASS = Class, COREPREPPROD = CorePreparProd, PIECECOMPLETENESS = PieceCompleteness, CORTEX = Cortex, CORTEXLOCATION = CortexLocation, PLATFORMTYPE = PlatformType, PLATFORMCORTEX = PlatformCortex, SCARCOUNT = ScarCount, THICKNESS = Thickness,
        MAXWIDTH = MaxWidth, LENGTH = Length, COREPLATFORM = CorePlatform, MAINFACECORTEX = MainFaceCortex, 
        MAINFACESCARCOUNT = MainFaceScarCount, COREABANDON = CoreAbandon, RETOUCHEDPIECETYOPOLOGY = RetouchedPieceTypology) %>% 
  mutate(RETOUCHEDPIECETYOPOLOGY = case_when(
    RETOUCHEDPIECETYOPOLOGY == "Right truncation on retouched piece" ~ "Truncation",
    RETOUCHEDPIECETYOPOLOGY == "Double backed bladelet" | RETOUCHEDPIECETYOPOLOGY == "Backed bladelet parcial" ~ "Backed bladelet",
    RETOUCHEDPIECETYOPOLOGY == "Dentilhado" ~ "Denticulate",
    RETOUCHEDPIECETYOPOLOGY == "Burin on truncation" ~ "Burin",
    TRUE ~ RETOUCHEDPIECETYOPOLOGY)) %>% 
  dplyr::select(ID, CLASS, COREPREPPROD, PIECECOMPLETENESS, CORTEX, CORTEXLOCATION, PLATFORMTYPE, PLATFORMCORTEX,
                SCARCOUNT, THICKNESS, MAXWIDTH, LENGTH, COREPLATFORM, MAINFACECORTEX, MAINFACESCARCOUNT, COREABANDON,
                RETOUCHEDPIECETYOPOLOGY)
gravettian_tech <- gravettian_tech %>% 
  mutate(RETOUCHEDPIECETYPOLOGY = case_when(
    RETOUCHEDPIECETYPOLOGY == "Splintered piece with notch" | RETOUCHEDPIECETYPOLOGY == "Splintered piece with lateral retouch" ~ "Splintered piece",
    RETOUCHEDPIECETYPOLOGY == "Retouched flake siret fragment" | RETOUCHEDPIECETYPOLOGY == "Retouched flake frag" ~ "Retouched flake fragment",
    RETOUCHEDPIECETYPOLOGY == "Retouched bladelet frag" ~ "Retouched bladelet fragment",
    RETOUCHEDPIECETYPOLOGY == "Multiple burin" | RETOUCHEDPIECETYPOLOGY == "Burin splintered piece" | RETOUCHEDPIECETYPOLOGY == "Burin on truncation" ~ "Burin",
    RETOUCHEDPIECETYPOLOGY == "Double backed bladelet" ~ "Backed bladelet",
    TRUE ~ RETOUCHEDPIECETYPOLOGY))

terrace_tech <- bind_rows(proto_tech_WIP, gravettian_tech)

terrace_RM_WIP <- terrace_RM %>%
  mutate(RMTYPE = case_when(
    RMTYPE == "Type 2" ~ "TYPE 2",
    RMTYPE == "Chalcedony" ~ "TYPE 1", # Altering first group names to final group names. Subtypes are maintained as before.
    RMTYPE == "TYPE 15" ~ "TYPE 3",
    RMTYPE == "TYPE 7" ~ "TYPE 4",
    RMTYPE == "TYPE 12" ~ "TYPE 5",
    RMTYPE == "TYPE 11" ~ "TYPE 6",
    RMTYPE == "TYPE 13" ~ "TYPE 7",
    RMTYPE == "TYPE 11F" ~ "TYPE 8",
    RMTYPE == "TYPE 13B" ~ "TYPE 9",
    RMTYPE == "Oolitic" ~ "TYPE 10",
    RMTYPE == "TYPE GB" ~ "TYPE 11",
    .default = RMTYPE)) %>% 
  select(AREA, ID, RMTYPE, SUBTYPE, WEIGHT, CORTEXTHICKNESS, CORTEXAPPEARANCE) %>%
  distinct(ID, .keep_all = TRUE)

context <- context %>%
  dplyr::rename("UNIT" = "Unit") %>% # Renaming first column from Unit to UNIT to use cols_to_concat
  unite(col='ID', cols_to_concat_terrace, sep="-", remove=FALSE) %>% # Unite UNIT with ID using cols_to_concat
  select(ID, Level, Spit, Code) %>% # Selecting important variables: ID refers to artifact ID; Level refers to archaeologycal layer; Spit refers to artificial levels within an archaeological layer; Code refers to type of artifact (e.g., LITHIC or BONE)
  distinct(ID, .keep_all = TRUE)

xyz <- xyz %>%
  unite(col='ID', cols_to_concat_terrace, sep="-", remove=FALSE) %>% # Unite UNIT with ID using cols_to_concat
  filter(Suffix == 0) %>% # To avoid several entries related to several points of the same artefact
  select(ID, X, Y, Z) %>% # Selecting important variables
  distinct(ID, .keep_all = TRUE)

field_data <- full_join(xyz, context, by = "ID") # Join field tables

terrace_tech_WIP <- left_join(terrace_tech, field_data, by = "ID") # Join field and RM type tables
terrace_dataset <- full_join(terrace_tech_WIP, terrace_RM_WIP, by = "ID") # Base 12690
terrace_dataset <- terrace_dataset %>%
  filter(!Level %in% c("3", "4", "4C", "4D")) %>%  # Dropping NAs on the technological class drops dataset to 1651
  drop_na(RMTYPE) %>%  # Dropping NAs on the RM database drops dataset to 3635
  drop_na(CLASS) # Dropping NAs on the technological class drops dataset to 1647
  
# Mutate level 5 to add phases: top of 5 (4E/5) and bottom of 5 (5), following Belmiro et al. (2020).
phase_dataset <- terrace_dataset %>% 
  mutate(Level = case_when(
    Level == 5 & Z > 24.1 ~ "4E/5",
    Level == 5 & Z < 24.1 ~ "5",
    Level == "7C" | Level == 8 ~ "7", # Collapsing samples from Level 7C and Level 8 into 7 to homogenise dataset
    TRUE ~ Level)) %>% # Adding technocomplex name based on the levels
  mutate(Technocomplex = case_when(
    Level == "4E" | Level == "4E/5" | Level == 5 ~ "Proto-Solutrean",
    Level == 6 | Level == 7 ~ "Gravettian"))

```


```{r}
#| label: shelter-solutrean-setup

cols_to_concat_shelter <- c("YEAR", "ID") # Concatenate YEAR and ID variables in one single variable (e.g. 05-500)

solutrean_WIP <- solutrean_JC %>% 
  filter(Layer %in% c("A", "B", "C")) %>% # Filtering Solutrean layers
  filter(Raw.material %in% c("Chert", "Chalcedony")) %>% # Filtering raw materials
  filter(Screen != "Crivo") %>%  # Filtering out all bucket lithics since they were not included in the RM analysis
  mutate_at("Date", str_replace, "20", "") %>% # Removing initial part of the year to match IDs
  dplyr::rename("YEAR" = "Date") %>% 
  tidyr::unite(col='ID', all_of(cols_to_concat_shelter), sep="-", remove=FALSE) # Concatenate YEAR and ID variables in one single variable (e.g. 05-500)


# In Cascalheira 2010 Retouched piece was not a category but a variable. To match the other DBs, this category was added
solutrean_WIP <- solutrean_WIP %>% 
  mutate(Class = case_when( # This alteration is being done on the original class column
    Retouched == "Yes" ~ "Retouched piece", # Whenever retouch is present, class is now Retouched piece
    TRUE ~ Class)) %>% 
  mutate(CLASS = case_when( # Creating a column for clustered classes, to match the other DBs
    Class == "Blade" | Class == "Bladelet" ~ "ElongBlank", # Cluster blades with bladelets. The distinction can come later through a mutate and the measurements
    Class == "Core front" | Class == "Core tablet" | Class == "Core trimming elem." | Class == "Crested piece" ~ "CorePrepProd", # Cluster all preparation products
    Class == "Fragment" ~ "Shatter",
    Class == "Flake distal frag." | Class == "Flake mesial frag." | Class == "Flake prox. frag." ~ "FlakeFrag", # Cluster flake frags into a single category
    Class == "Blade distal frag." | Class == "Blade mesial frag." | Class == "Blade prox. frag." 
    | Class == "Bladelet distal frag." | Class == "Bladelet mesial frag." | Class == "Bladelet prox. frag." ~ "ElongBlankFrag", # Cluster elong frags into a single category
    TRUE ~ Class)) %>% 
  mutate(COREPREPPROD = case_when( # Creating a column for the types of core preparation and maintenance products (COREPREPPROD)
    Class == "Core front" ~ "Core front",
    Class == "Core tablet" ~ "Core tablet",
    Class == "Core trimming elem." ~ "Core trim",
    Class == "Crested piece" ~ "Crested piece")) %>% 
  mutate(PIECECOMPLETENESS = case_when( # Creating a column for the portions present in flake or elongblank fragments to match other DBs
    Class == "Flake distal frag." | Class == "Blade distal frag." | Class == "Bladelet distal frag." ~ "Distal",
    Class == "Flake mesial frag." | Class == "Blade mesial frag." | Class == "Bladelet mesial frag." ~ "Mesial",
    Class == "Flake prox. frag." | Class == "Blade prox. frag." | Class == "Bladelet prox. frag." ~ "Proximal")) %>% 
  rename(PLATFORMTYPE = Platform_type) %>% # Renaming column
  mutate(PLATFORMTYPE = case_when( # Renaming variables to match correspondent variables in the other DBs
    PLATFORMTYPE == "Multifacetted" ~ "Faceted",
    PLATFORMTYPE == "Unfacetted" ~ "Plain",
    TRUE ~ PLATFORMTYPE)) %>%
  mutate(Type = case_when( # Translating retouched types from Zilhão 1997 to simplified typologies
    Type == "92b" ~ "Retouched piece fragment",
    Type == "92a" ~ "Retouched blank",
    Type == "76" ~ "Splintered piece",
    Type == "75" ~ "Denticulate",
    Type == "74" ~ "Notch",
    Type == "66" ~ "Retouched blade",
    Type == "61" ~ "Truncation",
    Type == "28" ~ "Burin",
    Type == "7" | Type == "5b" | Type == "4"  ~ "Endscrapper",
    TRUE ~ Type)) %>%
  rename(LEVEL = Layer, UNIT = Unit, CORTEX = Percent_of_cortex, CORTEXLOCATION = Loc_of_cortex, THICKNESS = Thickness,
         MAXWIDTH = Width, LENGTH = Length, COREPLATFORM = Core_platfform, SCARCOUNT = N_of_Scars, COREABANDON = Core_abandon,
         RETOUCHEDPIECETYPOLOGY = Type)  # Renaming column
  
solutrean_tech <- solutrean_WIP %>%
  dplyr::select(ID, CLASS, LEVEL, UNIT, COREPREPPROD, PIECECOMPLETENESS, CORTEX, CORTEXLOCATION, PLATFORMTYPE,
                SCARCOUNT, THICKNESS, MAXWIDTH, LENGTH, COREPLATFORM, COREABANDON, RETOUCHEDPIECETYPOLOGY)

solutrean_points <- solutrean_points %>% # Creating ID from 
  mutate_at("YEAR", str_replace, "20", "") %>% # Removing initial part of the year to match IDs
  tidyr::unite(col='ID', all_of(cols_to_concat_shelter), sep="-", remove=TRUE)
  
solutrean_tech <- bind_rows(solutrean_tech, solutrean_points) # Uniting technology dataset with points (2167 to 2228)

solutrean_tech <- solutrean_tech %>% # Originally 2228
  distinct(ID, .keep_all = TRUE) # Keeping individual IDs to avoid buckets (down to 1993)

shelter_RM_WIP <- shelter_RM %>%
  mutate_at("ID", str_replace, "VB", "") %>% # Remove "VB" from ID
  mutate(RMTYPE = case_when(
    RMTYPE == "Type 2" ~ "TYPE 2",
    RMTYPE == "Chalcedony" ~ "TYPE 1", # Altering first group names to final group names. Subtypes are maintained as before.
    RMTYPE == "TYPE 15" ~ "TYPE 3",
    RMTYPE == "TYPE 7" ~ "TYPE 4",
    RMTYPE == "TYPE 12" ~ "TYPE 5",
    RMTYPE == "TYPE 11" ~ "TYPE 6",
    RMTYPE == "TYPE 13" ~ "TYPE 7",
    RMTYPE == "TYPE 11F" ~ "TYPE 8",
    RMTYPE == "TYPE 13B" ~ "TYPE 9",
    RMTYPE == "Oolitic" ~ "TYPE 10",
    RMTYPE == "TYPE GB" ~ "TYPE 11",
    RMTYPE == "ND_A" ~ "INDET",
    .default = RMTYPE)) %>%
  select(AREA, ID, RMTYPE, SUBTYPE, WEIGHT, CORTEXTHICKNESS, CORTEXAPPEARANCE) %>%  # Select variables
  distinct(ID, .keep_all = TRUE) # Keeping individual IDs to avoid buckets

solutrean_dataset <- full_join(solutrean_tech, shelter_RM_WIP, by = "ID") # The complete dataset has 2818 (44 points) entries
solutrean_dataset <- solutrean_dataset %>%
  drop_na(RMTYPE) %>%   # Dropping NAs on the RM database drops dataset to 1514
  drop_na(CLASS) # Dropping NAs on the technological class drops dataset to 914
  
```

```{r}

library(ggplot2)
library(scales)

solutrean_dataset %>%
  filter(RMTYPE != "INDET") %>% 
  dplyr::group_by(LEVEL, RMTYPE) %>% 
  dplyr::summarise(n = n()) %>%
  mutate(percent = n/sum(n)) %>% 
  ggplot(aes(x = LEVEL, y = percent, fill = RMTYPE)) +
  geom_bar(stat = "identity", width = 0.5) +
  scale_y_continuous(labels = scales::percent_format()) +
  #theme_minimal() + #type of theme
  labs(y = "Count %") +# change labels
  guides(fill=guide_legend(title=NULL)) + # remove legend title
  geom_text(aes(label=ifelse(percent >= 0.05, paste0(sprintf("%.1f", percent*100),"%"),"")),
                position=position_stack(vjust=0.5), colour="white", size=3) +
  theme(axis.title.x = element_blank()) + 
  scale_fill_manual(values = c("#C74B50", "#D49B54", "#911F27", "#630A10",
                              "#FACE7F", "#8B6969", "#CDAF95", "#2B4F60",
                              "#996666", "#BDD2B6", "#798777", "#BDC7C9"))


```



Here is an example of inline code `r x` in the middle of a sentence.

# Discussion

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r}
#| label: colophon
#| cache: false

# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```
